Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-28 paper/main.tex
update text for extraction algorithm

% --------------------------------------------------------------------------------------------------
\subsection{Visualization}
\label{subsection:visualization}
To visualize the spatial configuration of the results we employ
\texttt{pyvista}~\cite{sullivan2019pyvista}, which provides
high-level wrappers around the popular VTK library~\cite{vtkBook,Sullivan2019}.
The initial phase of the visualization process entails the generation of three-dimensional meshes
for each bacterium as an intermediate representation (see
Figure~\ref{fig:progression-image-generation}).
We insert spheres matching the thickness of the rod $R$ at each vertex $\vec{x}_i$ and connect
adjacent vertices $\vec{x}_i,\vec{x}_{i+1}$ with a cylinder, of radius $r$ which is positioned
such that its core is identical to the connecting line segment between the two vertices.
Finally, we combine the meshes of all of these objects to obtain a single mesh of one bacterium.
A render including smoothened meshes and external light sources of this configuration can be seen in
Figure~\ref{fig:progression-image-generation} (B).

@@

Figure~\ref{fig:benchmarking-extraction-algorithm} (D-H) shows snapshots of the synthetically
generated cell masks.
The initial configuration consists of 4 bacteria which grow and ultimately divide thus expanding
the filled up area.
Meanwhile figure~\ref{fig:benchmarking-extraction-algorithm} (I) shows a zoomed-in version of
snapshot (G), highlighting the direct comparison of the exact (dark cross) and extracted vertices.
Visual comparison of this time series indicates strong alignment between exact and extracted
results.
Upon closer inspection with figure~\ref{fig:benchmarking-extraction-algorithm} (I), we can see
that small differences start to manifest, especially at the tip of the rods in more crowded regions.

In order to quantitatively analyze this disparity we use both data to calculate the average rod
length of the bacteria (Figure~\ref{fig:benchmarking-extraction-algorithm} (J)) over the course of
the simulated time interval.
The results show good agreement with larger discrepancies directly after division events.
In addition figure~\ref{fig:benchmarking-extraction-algorithm} (K) displays individual displacements
of exact and extracted vertices which have been reoriented with respect to the direction of the rod.
Values closer to the center represent better agreement.
The distribution of these displacements can be well described by a normal distribution with mean
$\mu$ and variance $\sigma$.
These parameters can be obtained by fitting the distribution to the displacements at every time
point which is displayed in Figure~\ref{fig:benchmarking-extraction-algorithm}.
Here we can also see that the overall variance in the displacements increases directly after
division events.
However, the center of the distributions remain concentrated near zero without significant outliers
which demonstrates the robustness of our method in these densely packed scenarios.
Further details are given in the supplementary materials
(section~\ref{sec:supplement-imaging-based-comparison-algorithm-details}).

\paragraph{Determining the underlying Error Model}

@@

        {docs/source/_static/fitting-methods/extract_positions-007800-cropped.pdf}%

@@

        (I) Average rod length of the exact and extracted data.

@@

        (J) Displacement of individual vertices parallel and orthogonal to the rod.

@@

        (K) Mean $\mu_{p,o}$ and variance $\sigma_{p,o}$ of fitted normal distribution for
        displacements parallel and orthogonal to the rod displayed over the duration of the
        simulation.

>>>>

% --------------------------------------------------------------------------------------------------
\subsection{Visualization}
\label{subsection:visualization}
Our visualization routine relies on just the spatial representation in the form of vertices and the
thickness of the rod which is assumed to be uniform along the rod.
To visualize the spatial configuration of the results we employ
\texttt{pyvista}~\cite{sullivan2019pyvista}, which provides
high-level wrappers around the popular VTK library~\cite{vtkBook,Sullivan2019}.
The initial phase of the visualization process entails the generation of three-dimensional meshes
for each bacterium as an intermediate representation (see
Figure~\ref{fig:progression-image-generation}).
We insert spheres matching the thickness of the rod $R$ at each vertex $\vec{x}_i$ and connect
adjacent vertices $\vec{x}_i,\vec{x}_{i+1}$ with a cylinder, of radius $r$ which is positioned
such that its core is identical to the connecting line segment between the two vertices.
Finally, we combine the meshes of all of these objects to obtain a single mesh of one bacterium.
A render including smoothened meshes and external light sources of this configuration can be seen in
Figure~\ref{fig:progression-image-generation} (B).

@@

Figure~\ref{fig:benchmarking-extraction-algorithm} (D-H) shows snapshots of the synthetically
generated cell masks.
The initial configuration consists of four bacteria which grow and divide thus expanding the
occupied area.
Meanwhile figure~\ref{fig:benchmarking-extraction-algorithm} (I) shows a zoomed-in version of
snapshot (G), highlighting the direct comparison of the exact (white cross) and extracted vertices.
Visual comparison of this time series indicates strong alignment between exact and extracted
results.
Upon closer inspection with figure~\ref{fig:benchmarking-extraction-algorithm} (I), we can see
that small differences start to manifest, especially at the tip of the rods in more crowded regions.

In order to quantitatively analyze this disparity we use both data to calculate the average rod
length of the bacteria (Figure~\ref{fig:benchmarking-extraction-algorithm} (J)) over the course of
the simulated time interval.
The results show good agreement with larger discrepancies directly after division events.
In addition figure~\ref{fig:benchmarking-extraction-algorithm} (K) displays individual displacements
of exact and extracted vertices which have been reoriented with respect to the direction of the rod.
Values closer to the center represent better agreement.
The distribution of these displacements can be well described by a normal distribution with mean
$\mu$ and variance $\sigma$.
These parameters can be obtained by fitting the distribution to the displacements at every time
point.
The parameters resulting from this fit are displayed in
Figure~\ref{fig:benchmarking-extraction-algorithm} (L), where the overall variance (filled area) in
the displacements significantly increases directly after division events.
However, the centers of the distributions (solid curves) remain concentrated near zero without
significant outliers which demonstrates the robustness of our method in these densely packed
scenarios.
These observations provide us with enough information to construct an error model for the extraction
algorithm.
Further details are given in the supplementary materials
(section~\ref{sec:supplement-imaging-based-comparison-algorithm-details}).

@@

        {docs/source/_static/fitting-methods/extract_positions-009800-cropped.pdf}%

@@

        (I) Zoomed-in version of (H).
        (J) Average rod length of the exact and extracted data.

@@

        (K) Displacement of individual vertices parallel and orthogonal to the rod.

@@

        (L) Mean $\mu_{p,o}$ (curve) and variance $\sigma_{p,o}$ (filled space) of fitted normal
        distribution for displacements parallel and orthogonal to the rod displayed over the
        duration of the simulation.
